Self-tunable approximated explicit MPC: Heat exchanger implementation and analysys

### 2023-02-03 paper/elsarticle/main_LOCAL_471.tex
relative energy

	The current crisis of energy resources emphasizes the long-term goal of achieving sustainable industrial production and optimal energy utilization. Moreover, minimizing the energy utilization directly reduces the corresponding CO$_{2}$ emissions. Therefore, sustainable industrial production is focused on the wide implementation of advanced control methods~\cite{MN20}. Recent survey on the applied thermal engineering focused on the energy saving and pollution reduction from the industrial perspective is provided in~\cite{YV16}.

	The heat exchangers in their numerous variants are integrated into the many industrial plants as the heat transfer represents the crucial phenomena for all thermal energy applications~\cite{KV18}. Simultaneously, the heat exchangers are energy-demanding and resources-intensive plants. Moreover, from the control viewpoint, the controller design taking into account the nonlinear and asymmetric behaviour of the heat exchangers is a challenging task, see~\cite{RL20}. Although the conventional and widely-used proportional–integral–derivative (PID) controllers are robust and easy to tune, their control performance may not be sufficient when it comes to energy savings. Various extensions built above the well-tuned PID controller were developed to compensate for the nonlinear and asymmetric behavior, often affected by the additional impact of the uncertain parameters originated, e.g., in the fouling~\cite{MT19}. Such widely-used control strategies including, e.g., the robust control~\cite{WY18}, the gain-scheduling, self-tuning, adaptive and optimal control, see~\cite{MF08} for the rigorous mathematical modeling and controller design methods in general, and~\cite{Liptak} for the controller design tailored for the process control engineers.

	One of the promising control strategies addressing all these issues in an optimal way came with the formation and tuning of the model predictive control (MPC), e.g., see~\cite{Morari_MPC}. MPC provides optimal control input based on the minimization of a specified cost function while considering a model of the system. Compared to linear quadratic controllers (LQR)~\cite{LQR}, model predictive control also includes constraints on the control input or process variables~\cite{Maciejowski_MPC}, and additional saturation is not necessary. Moreover, as the optimization problem is solved in every control step, MPC represents a receding horizon control policy~\cite{receding_horizon}, having a significant benefit mainly in the terms of disturbance rejection. The model predictive control was intensively investigated in connection with heat exchangers. In~\cite{Vinaya_HE_MPC}, the authors developed a model predictive control for a shell and tube heat exchanger. In~\cite{Oravec_energy}, the control performance of the robust MPC implemented for a plate heat exchanger with soft constraints was analyzed. Four robust control strategies were presented and compared in~\cite{Oravec_HE_ATE}. A two-level control structure was applied on a heat exchanger network in~\cite{Gonzales_HE_MPC}, where the low level of control was ensured by MPC and the high level by a supervisory online optimizer. The fast nonlinear MPC was designed to optimize the waste heat recovery~\cite{WC19}.  The multi-layer control designed in~\cite{DZ18} designed the MPC in the leader loop to optimize the thermal response to the improved control performance.

@@

	\section{Theoretical backgrounds}
	\label{sec:preliminaries}

@@

To analyze the control more objectively, energy consumption was also investigated. More specifically, the power used to heat the heating medium was measured and summarized for the whole control but for every controller setup, respectively. The energy consumption $E$ necessary for every control can be seen in Table~\ref{tab:energy}.

\begin{table}[h!]
	\begin{center}
		\caption{Energy consumption evaluation.}
		\label{tab:energy}
		\begin{tabular}{c|c} 
			$Q_\mathrm{y}$ & E\,[kJ]   \\
			\hline
			1000 &  883  \\ 
			% \hline
			100 & 938   \\ 
			% \hline
			self-tuned & 921   
		\end{tabular}
	\end{center}
\end{table}

@@

	This paper deals with the experimental implementation of the self-tunable approximated explicit model predictive control and provides a strategy for an effective self-tuning controller design. Based on the current reference value, the tuning parameter is scaled using linear interpolation. The previously published work related to the self-tunable explicit MPC suggested tuning based on the distance of the reference value from the system steady-state value. This paper presents a novel idea of tuning based on the size of reference step change. The self-tuning algorithm aims to compensate for the nonlinear behavior of the controlled system. The self-tuning parameter is updated when the reference changes. It is calculated as the ratio between the size of the reference change and the maximal possible size of the reference change, which is specified before operation. 
	Another novel contribution addresses the challenging control problem of asymmetric system behavior by splitting the interval of tuning parameter into two ranges, while both intervals are assigned to different operating conditions. The proposed method is implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. The asymmetry makes the plant a suitable candidate for splitting the interval of the tuning parameter. The decision criterium is negativity or positivity of reference change. When the reference changed upwards, the control input was tuned in the first part of the interval and approached the boundary controller associated with the lower bound on the penalty matrix. On the contrary, when the reference changed downwards, the control input was tuned to approach the control input from the boundary controller with the upper bound on the penalty matrix. To investigate the control results properly, the control performance was evaluated also quantitatively. Compared to the conventional control strategy handling just a single controller, the self-tunable control approach decreased the integral squared control error, maximal overshoots/undershoots, and settling time. In addition, the energy consumption corresponding to every controller was evaluated as well. The value of consumed energy was between the two boundary controllers, i.e., the energy consumption did not deteriorate in general. With saving energy compared to one of the boundary controllers, the control performance was significantly improved.

>>>>

	The current crisis of energy resources emphasizes the long-term goal of achieving sustainable industrial production and optimal energy utilization. Moreover, minimizing the energy utilization directly reduces the corresponding CO$_{2}$ emissions. Therefore, sustainable industrial production is focused on the wide implementation of advanced control methods~\cite{MN20}.

	The heat exchangers, in their numerous variants integrated into many industrial plants, represent energy-demanding plants. From the control viewpoint, the controller design taking into account the nonlinear and asymmetric behavior of the heat exchangers is a challenging task, see~\cite{RL20}. Although the conventional and widely-used proportional–integral–derivative (PID) controllers are robust and easy to tune, their control performance may not be sufficient when it comes to energy savings. Various extensions built above the well-tuned PID controller were developed to compensate for the nonlinear and asymmetric behavior, often affected by the additional impact of the uncertain parameters originated, e.g., in the fouling~\cite{MT19}. Such widely-used control strategies include, e.g., the robust control~\cite{DO15,WY18}, the gain-scheduling, multi-zone control, see~\cite{MF08}, and references therein.

	TODO: nepouzivat nasobne referencie ako [4,5] vyssie

	One of the promising control strategies addressing all these issues in an optimal way came with the formation and tuning of the model predictive control (MPC), e.g., see~\cite{Morari_MPC}. MPC provides optimal control input based on the minimization of a specified cost function while considering a model of the system. Compared to linear quadratic controllers (LQR)~\cite{LQR}, model predictive control also includes constraints on the control input or process variables~\cite{Maciejowski_MPC}, and additional saturation is not necessary. Moreover, as the optimization problem is solved in every control step, MPC represents a receding horizon control policy~\cite{receding_horizon}, having a significant benefit mainly in the terms of disturbance rejection. The model predictive control was intensively investigated in connection with heat exchangers. In~\cite{Vinaya_HE_MPC}, the authors developed a model predictive control for a shell and tube heat exchanger. In~\cite{Oravec_energy}, the control performance of the robust MPC implemented for a plate heat exchanger with soft constraints was analyzed. Four robust control strategies were presented and compared in~\cite{Oravec_HE_ATE}. A two-level control structure was applied on a heat exchanger network in~\cite{Gonzales_HE_MPC}, where the low level of control was ensured by MPC and the high level by a supervisory online optimizer.

@@

	\section{Theoretical backgrounds}
	% \section{Preliminaries}
	\label{sec:preliminaries}

@@

To analyze the control more objectively, energy consumption was also investigated. More specifically, the power used to heat the heating medium was measured and summarized for the whole control but for every controller setup, respectively. The energy consumption $E$ necessary for every control as well as the relative improvement subject to every boundary controller can be seen in Table~\ref{tab:energy}.

\begin{table}[h!]
	\begin{center}
		\caption{Energy consumption evaluation.}
		\label{tab:energy}
		\begin{tabular}{c|c|c} 
			$Q_\mathrm{y}$ & E\,[kJ] & $\delta E\,[\%]$   \\
			\hline
			1000 &  883 & -4\\ 
			% \hline
			100 & 938 & 2 \\ 
			% \hline
			self-tuned & 921 & -   
		\end{tabular}
	\end{center}
\end{table}

@@

	This paper deals with the experimental implementation of the self-tunable approximated explicit model predictive control and provides a strategy for an effective self-tuning controller design. Based on the current reference value, the tuning parameter is scaled using linear interpolation. The previously published work related to the self-tunable explicit MPC suggested tuning based on the distance of the reference value from the system steady-state value. This paper presents a novel idea of tuning based on the size of reference step change. The self-tuning algorithm aims to compensate for the nonlinear behavior of the controlled system. The self-tuning parameter is updated when the reference changes. It is calculated as the ratio between the size of the reference change and the maximal possible size of the reference change, which is specified before operation. 
	Another novel contribution addresses the challenging control problem of asymmetric system behavior by splitting the interval of tuning parameter into two ranges, while both intervals are assigned to different operating conditions. The proposed method is implemented on a laboratory heat exchanger with nonlinear and asymmetric behavior. The asymmetry makes the plant a suitable candidate for splitting the interval of the tuning parameter. The decision criterium is negativity or positivity of reference change. When the reference changed upwards, the control input was tuned in the first part of the interval and approached the boundary controller associated with the lower bound on the penalty matrix. On the contrary, when the reference changed downwards, the control input was tuned to approach the control input from the boundary controller with the upper bound on the penalty matrix. To investigate the control results properly, the control performance was evaluated also quantitatively. Compared to the conventional control strategy handling just a single controller, the self-tunable control approach decreased the integral squared control error up to 32\%, maximal overshoots/undershoots up to 18\%, and settling time up to 24\%. In addition, the energy consumption corresponding to every controller was evaluated as well. The value of consumed energy was between the two boundary controllers, i.e., the energy consumption did not deteriorate in general. With saving the energy compared to one of the boundary controllers, the control performance was significantly improved.
